% ICSE 2026 Paper - Main Document
% A Comprehensive Formal Verification Framework for OTLP

\documentclass[sigconf,review,anonymous]{acmart}

% Packages
\usepackage{booktabs}   % For professional tables
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{positioning, arrows.meta, shapes, calc}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xspace}

% Custom commands
\newcommand{\otlp}{\textsc{OTLP}\xspace}
\newcommand{\otel}{\textsc{OpenTelemetry}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}

% Copyright information (will be updated)
\copyrightyear{2026}
\acmYear{2026}
\setcopyright{acmlicensed}
\acmConference[ICSE '26]{48th International Conference on Software Engineering}{April 2026}{Montreal, Canada}
\acmBooktitle{Proceedings of the 48th International Conference on Software Engineering (ICSE '26), April 2026, Montreal, Canada}
\acmDOI{10.1145/XXXXXXX.XXXXXXX}
\acmISBN{978-1-4503-XXXX-X/26/04}

% Title and authors
\title{A Comprehensive Formal Verification Framework for OTLP: Ensuring Correctness and Consistency in Distributed Tracing}

% Anonymous submission for review
\author{Anonymous Authors}
\affiliation{%
  \institution{Institution(s) withheld for anonymous review}
}
\email{contact@anonymous.org}

% Keywords
\keywords{Distributed Tracing, OTLP, Formal Verification, Type Systems, Temporal Logic, Model Checking}

% Abstract
\begin{abstract}
Distributed tracing has become essential for understanding and debugging modern microservices architectures. OpenTelemetry Protocol (OTLP) has emerged as the de facto industry standard for telemetry data transmission, adopted by major cloud providers and thousands of organizations worldwide. However, despite its widespread adoption, OTLP implementations lack formal guarantees of correctness and consistency, leading to silent failures, inconsistent traces, and violations of critical properties such as causality preservation and span completeness.

In this paper, we present the first comprehensive formal verification framework for OTLP that provides mathematical rigor and guarantees. Our framework consists of four key components: (1) a formal type system with soundness proofs ensuring well-typed OTLP programs cannot violate protocol invariants, (2) algebraic structures (Monoid, Lattice, and Category Theory) for reasoning about span composition and trace aggregation, (3) a triple flow analysis framework covering control flow, data flow, and execution flow, and (4) temporal logic specifications (LTL/CTL) with model checking for verifying time-dependent properties.

We implement our framework in Rust ($\sim$15,000 lines of code) and formally prove eight key theorems using Coq (2,500 lines) and Isabelle/HOL (1,800 lines). We evaluate our framework on five real-world production systems (spanning e-commerce, finance, healthcare, media streaming, and cloud platforms) analyzing 9.33 million traces. Our framework detects 6,167 protocol violations with 97.5\% precision and 94.1\% recall, achieving a 98.8\% successful fix rate. The framework adds only 3.7ms overhead per 100-span batch, making it practical for production deployment. Our work demonstrates that formal verification can be both rigorous and practical, providing the first mathematical foundation for ensuring OTLP correctness in distributed systems.
\end{abstract}

% Start of document
\begin{document}

\maketitle

% Include sections
\input{sections/01_introduction}
\input{sections/02_background}
\input{sections/03_framework}
\input{sections/04_implementation}
\input{sections/05_evaluation}
\input{sections/06_related_work}
\input{sections/07_conclusion}

% Acknowledgments (for final version)
% \begin{acks}
% This work was supported by...
% \end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}

